### 2009-11-14 docs/marcello-dissertation/thesis/chapter1-introduction.tex
Updating the thesis and added chapters 2-5. Still revisions are needed in section 4 and 5, adjusting images and tables. Images are in the eps format.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@642 559edde2-a4d4-44e0-96fc-4035e40d7396

In order to provide access to the collected sensor data, different approaches
can be taken to interrogate sensor devices. In general, the query process can
take place on the network nodes themselves, as it is called in-network, or in 
a centralized data sink. The former is generally used when the data is
collected at the sensor node on its local storage, being shown by \cite{sn-storage01}
\cite{sn-storage03} \cite{sn-storage04} that this strategy helps mitigating energy
consumption. However, this strategy is used when data archival is not the
primary reason, given the limited resources on the sensor devices. For this
reason, the the latter strategy has been used when the collected data is
primarily archived in a centralized database system for later reuse
\cite{sn-storage02}.

When it comes to data model implementation, the relational data model
\cite{relational-model} is one of the most used approaches to represent data in
sensor network databases in any of the query processing strategies, and in most cases 
with a modified version of the SQL language \cite{sn-db-newop01}
\cite{sn-db-newop02}. However, the design of the database system must be done
with prior knowledge of the sensor types when  choosing the relational data
model, as the database design must be normalized \cite{db-normalization} for the entities
already identified for the system. For instance, the introduction of a new
sensor device might represent a potential change on the database structure by
seeking a new normalized version of the database structure. In addition to
challenges of maintaining the database schema normalized, the use of Data
Provenance approaches in sensor network data \cite{sn-provenance01} usually
addresses the problem related with the lineage of data  and its full history.
However, once the data model is in place to receive the collected data, one of
the most important non-functional requirements in sensor networks is
scalability of the persistence storage. Since an operational sensor network
can produce huge amounts of data, it must be able to cope with the increase
and data load during data gathering, and yet provide users the same
performance during data retrieval by using different strategies such a
data-centric storage \cite{sn-storage03}.

In summary, this dissertation discusses the solution of a centralized
data-centric persistence storage component for the NetBEAMS \cite{netbeams09},
Data Sensor Platform (DSP), an open-source \cite{open-source} implementation of
a sensor network that primarily focused on environmental data produced by NetBEAMS. 
Different challenges can be related to the DSP in the context of sensor 
networks: data model, data persistence organization and strategy, and user data
access and distribution. First, NetBEAMS needs to provide means to safely 
represent sensor devices' properties without the need of restructuring any 
existing data model, taking into account the use cases from the main users. 
For example, a marine biologist may query the the persistence storage regarding
the temperature of a specific geographic region at the San Francisco Bay at a 
given time frame. Second, since a sensor network may experiment grown in terms 
of number of sensors, and consequently the number of data produced, the
persistence  storage must be able to reasonably scale and be well structured
in terms of storage space with the least service interruption. For instance,
if the persistence storage is close to reach its storage capacity, what
should the network administrator do? Last, but not least, the means to access 
the data for reuse is the most important features of the system, taking into 
account the different data formats used by the main users of the system. How 
to export specific collected data to a spreadsheet or archive files to be 
published or shared with other users? For these, and many other reasons, the 
persistence storage problem addressed by this work can be seen as a very 
challenging experience to solve.

In view of the fact that the persistence storage of a sensor network imposes
many different challenges, and taking into account the case study NetBEAMS, this
work's main strategy focused analyzing the different specifications of data
persistence in the literature in order to find new types of mechanisms to
provide data persistence for sensor network data. In this way, the kernel of
this dissertation is focused on providing solutions to the real-world problem
face by the Romberg Tiburon Center and its use of NetBEAMS. Furthermore,
the evaluation of different technologies is discussed in order to provide a
loosely-coupled component-based solution based on the infrastructure of the 
DSP Platform. For the reason of providing a 'lighter' data model, a subtype of
the Key-Value Pair \cite{db-kvp01} data model was chosen to easily capture the
description of any sensor device's properties, along with the use of time
series data models \cite{sn-provenance01} for time snapshots. In that way, data
modeling process does not represent a new refactoring each time a new sensor 
type is added or removed from the persistence system, giving users means to 
retrieve data accordingly. Similarly, in contemplation of scalability, this 
work proposes the use of an increasingly popular infrastructure technique called
Database Sharding \cite{db-shard01} \cite{db-shard02}, which takes the approach
of Data-Centric storage \cite{sn-storage03} on in-network query processing in a
centralized database system. In order to provide such mechanism, this work
evaluates the use of shard keys based on the geographic origin of the collected
data as cited to be one of the options when working with Provenance data in 
sensor networks \cite{sn-provenance01}. As a result, the data persistence system
component was conceived using the plug-and-play capability from the DSP 
platform without doing  any changes to the current DSP Platform, providing an
scalable solution that can adjust according to any data load increase. Finally,
the selected technology provides different data access mechanisms, which
directly addressed problems related to data access and Information Fusion 
\cite{sn-info-fusion}, providing export-import capabilities and visualization through
the availability of APIs written in different popular programming languages.

>>>>

In order to provide access to the collected sensor data, different approaches
can be taken to interrogate sensor devices indirectly by accessing its data in
places called network sink. These data nodes can be located in the network or 
in a centralized host. In this way, the query process can take place on the
network nodes themselves or on the centralized data sink. The former is
generally used when the data is collected at the sensor node on its local
storage, being shown by \cite{sn-storage01} \cite{sn-storage03}
\cite{sn-storage04} that this strategy helps mitigating energy consumption.
However, this strategy is used when data archival is not the primary reason,
given the limited resources on the sensor devices. For this reason, the the
latter strategy has been used when the collected data is primarily archived in
a centralized database system for later reuse \cite{sn-storage02}.

When it comes to data model implementation, the relational data model
\cite{relational-model} is one of the most used approaches to represent data in
sensor network databases in any of the query processing strategies, and in most cases 
with a modified version of the SQL language \cite{sn-db-newop}. However, the
design of the database system must be done with prior knowledge of the sensor
types when choosing the relational data model, as the database design must be
normalized \cite{db-normalization} for the entities already identified for the
system. For instance, the introduction of a new sensor device might represent
a potential change on the database structure by seeking a new normalized
version of the database structure. In addition to challenges of maintaining
the database schema normalized, the use of Data Provenance approaches in
sensor network data \cite{sn-provenance} usually addresses the problem related
with the lineage of data  and its full history. However, once the data model
is chosen and implemented, one of the most important non-functional
requirements in sensor networks is scalability of the persistence storage.
Since an operational sensor network can produce huge amounts of data, it must
be able to cope with the increase and data load during data gathering, and yet
provide users the same performance during data retrieval by using different
strategies such a data-centric storage \cite{sn-storage03}.

In summary, this dissertation discusses the solution of a
centralized data-centric persistence storage component for the NetBEAMS
\cite{netbeams2009}, an implementation of a sensor network that primarily
focused on environmental data. Different challenges can be related to the
selection of a technology that handles data persistence in the context of sensor
networks: the network infrastructure, the nature and location of the collected
data, the chosen data model, among others. First, NetBEAMS needs to provide
means to safely represent sensor devices' properties without the need of
restructuring any existing data model, taking into account the use cases from
the main users. For example, a marine biologist may query the the persistence
storage regarding the temperature of a specific geographic region at the San
Francisco Bay at a given time frame. Second, since a sensor network may
experiment grown in terms of number of sensors, and consequently the number of
data produced, the persistence  storage must be able to reasonably scale and
be well structured in terms of storage space with the least service
interruption. For instance, if the persistence storage is close to reach its
storage capacity, what should the network administrator do? Last, but not
least, the means to access the data for reuse is the most important features
of the system, taking into account the different data formats used by the main
users of the system. How to export specific collected data to a spreadsheet or
archive files to be published or shared with other users? For these, and many
other reasons, the persistence storage problem addressed by this work can be
seen as a very challenging experience to solve.

In view of the fact that the persistence storage of a sensor network imposes
many different challenges, and taking into account the case study NetBEAMS, this
work's main strategy focused analyzing the different specifications of data
persistence in the literature in order to find new types of mechanisms to
provide data persistence for sensor network data. In this way, the kernel of
this dissertation is focused on providing solutions to the real-world problem
face by the Romberg Tiburon Center and its use of NetBEAMS. Furthermore,
the evaluation of different technologies is discussed in order to provide a
loosely-coupled component-based solution based on the infrastructure of the 
DSP Platform. For the reason of providing a 'lighter' data model, a subtype of
the Key-Value Pair \cite{db-kvp} data model was chosen to easily capture the
description of any sensor device's properties, along with the use of time
series data models \cite{sn-provenance} for time snapshots. In that way, data
modeling process does not represent a new refactoring each time a new sensor 
type is added or removed from the persistence system, giving users means to 
retrieve data accordingly. Similarly, in contemplation of scalability and the
use of Data-Centric approach \cite{sn-storage03}, this work proposes the use of
Database Sharding \cite{db-shard01} \cite{db-shard02}, on a centralized
database system.

In order to provide such mechanism, this work first takes a look at the related
literature to study the properties of sensor networks, in the context of its
produced data and infrastructure for storage and query processing. Then, a
summarized list of taxonomies is proposed to be the basis of an empirical  the
use of shard keys based on the geographic origin of the collected data as cited to be one of the options when working with Provenance data in sensor networks \cite{sn-provenance}. As a result, the data persistence system component was conceived using the plug-and-play capability from the DSP platform without doing  any changes to the current DSP Platform, providing an scalable solution that can adjust according to any data load increase. Finally, the selected technology provides different data access
mechanisms, which directly addressed problems related to data access and
Information Fusion \cite{sn-info-fusion}, providing export-import capabilities
and visualization through the availability of APIs written in different
popular programming languages.

### 2009-11-14 docs/marcello-dissertation/thesis/chapter2-state-of-the-art.tex
Updating the thesis and added chapters 2-5. Still revisions are needed in section 4 and 5, adjusting images and tables. Images are in the eps format.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@642 559edde2-a4d4-44e0-96fc-4035e40d7396

\bibliography{thesis}  % uses the references stored in Chapter1Radar.bib

@@

\section{Sensor Networks}

@@

\section{Data Persistence in Sensor Networks}

@@

At first glance, the raw data serves as the primary product of the sensor
network. However, the first consideration that must take place is regarding the
purpose of the collected data, as \cite{sn-provenance01} explains
differentiates between two main use of the sensor data:

@@

While real-time data stream are used and can potentially be discarded, the
archival data needs to be stored in a secondary memory. Nonetheless, given the
nature of the sensor device specification, the low-level data produced
by a given sensor device might not contain necessary information regarding the
identity of the data itself. The use of Data Provenance techniques is described
by \cite{sn-provenance01} as an important step to provide the collected data 
specific metadata regarding its location and lineage.

@@

\subsubsection{Providing Identity to the Collected Sensor Data}

Most of sensor networks implementations use the collected data from the sensor
devices after they observe the environment. \cite{sn-provenance01} showed that
collected data used for archival purposes are usually missing important
features. First, the device's sampling are usually in raw format or without
obvious names. In order to provide search over the collected sensor data, the
data needs first to be described and indexed, and this process can be related
to Data Provenance \cite{db-provenance}.

An example of Data Provenance application is the use of aggregation of data
over time to estimate the changing effects of the weather in a given region.
For this reason, the collected data needs descriptions or annotations that can
give enough information to track changes, that is, the history of how and when
a given data has changed. The following should be the foundation of a
provenance-aware collected data according to \cite{sn-provenance01}:

@@

\subsection{Query Processing for Storage Locations}

@@

\begin{itemize}
  \item {In-Network Query Processing}: sensor nodes that provides real-time
  data stream usually replies to requests to properties. Likewise, when the
  data is used for archival purpose, the data can be located at a network node
  specialized in storing the neighbors collected data. In this way, when a
  query is issued to the sensor network, the query processor will
  collectively verify every storage node for the collected data.
  \item {Centralized Query Processing}: a main network sink receives the
  collected data from the peer nodes. In this case, the query processor queries
  the solo centralized node for the overall values of the participating sensor
  devices.
\end{itemize}

There are pros and cons related to both of the approaches of query processing,
which are also related to the infrastructure, sensor network nature and 
requirements. The main goal of the in-network query processing is to decrease
the amount of data transmitted from the sensor devices to the final data sink, as
sensor devices are usually limited by constraints such as energy utilization 
\cite{sn-storage03}. Along with this strategy, the literature describes
different proposals that aims at upgrading the Structured Query Language (SQL)
adding location-sensitive clauses to the language \cite{sn-db-newop01}
\cite{sn-db-newop02}. However, this main problem related to this approach is
that it may potentially flood all nodes with query messages \cite{sn-storage04}.

@@

A centralized network query processing processes data in a network node
specified as the network sink. The network sink usually is a network
node with a database server with the collection of data delivered by
all the sensor devices from the network. As a consequence of this
many-to-one communication strategy \cite{sn-storage02}, the query process does
not affect any of the sensor nodes in the network and, thus, maximizes the
sensor device utilization. Moreover, the use of the data for historical
purposes is easier to manage and give access to the primary end users of the
data, following the process of data processing, compression, aggregation, etc
\cite{sn-db02}.

@@

\section{SF-BEAMS: a Marine Sensor Network for Water Quality Monitoring}

The San Francisco Bay Environmental Assessment and Monitoring Station,
or SF-BEAMS, is a marine sensor network operated by the Romberg Tiburon Center
(RTC). This section details the SF-BEAMS sensor network located at
the Tiburon Island, California, and then presents the details of NetBEAMS, as
it is part of the case study of this work.

\subsection{The SF-BEAMS Infrastructure}

The RTC's main focuses are in the study of complex marine and estuarine
environments using the SF-BEAMS sensors deployed offshore of its pier located
on the banks of the San Francisco Bay at the Tiburon Island, California. In
this way, the project is responsible for providing data for water quality
monitoring, as well as weather and surface conditions.

In general, the SF-BEAMS network infrastructure contains heterogeneous wired and
wireless devices standing on pilons docked in the pier, where each of them is
responsible for observing different conditions of the area, as well as having
its own mechanisms for internal storage for the collected data. In this way,
data can be directly transferred to the labs via ethernet cables or collected
manually download by a staff's laptop computer. Figure \ref{fig:sf-beams}
shows a picture taken from the SF-BEAMS web-cam placed in-site.

\begin{figure}
  \centering
    \includegraphics{../diagrams/cam_image-oct15}
  \caption{Picture of the SF-BEAMS Network, at the RTC pier. Tiburon, CA.}
  \label{fig:sf-beams}
\end{figure}

\subsection{SF-BEAMS produced data and collection process}
\label{sec:sfbeams}

As described in section \ref{sec:sn-infrastructure}, sensor devices generates
its observed data based on properties of measurements defined by its
manufacturer. SF-BEAMS is a single-hop star sensor network containing
different instances of sensors, including the "YSI 6600 ESD V2", as seen on
figure \ref{fig:ysi-device}, which is a powerful water quality monitoring
device that produces around 52 bytes (13x4 Bytes) on a single real-time data
stream as shown on table \ref{tab:ysi-data-stream}.

\begin{figure}
  \centering
  \includegraphics{../diagrams/ysi-device}
  \caption{Picture of the case study sonde device: YSI 6600 ESD V2}
  \label{fig:ysi-device}
\end{figure}

\begin{table}
    \caption{YSI Data Stream}
    \begin{center}
        \begin{tabular}{lr}
  12.20    192    179 5588.40   0.09   0.084   0.059  7.98   -79.6   99.5  8.83 
  0.4     8.7
        \end{tabular}
    \end{center}
    \label{tab:ysi-data-stream}
\end{table}

In order to have access to the observed data, the network staff downloads
the data using one of the device's connection such as the RS-232 serial
connector \cite{rs232}. Then, the data is transferred to the RTC labs for
archival, where it is described, indexed and distributed to its main users.
According to its staff members, in May 2009 the infrastructure of the SF-BEAMS
included the following was organized as follows:

\begin{itemize}
  \item 5 YSI sondes in operation at the RTC site Pier in Tiburon Island, San
  Francisco Bay;
  \item Each YSI sonde produces 52 Bytes for each its observations;
  \item The sampling frequency rate could be configured in ranges of 1, 6 or
  15 minutes, depending on the staff's speficiation. 
\end{itemize}

Considering the infrastructure previously described, the size of the data
produced by the YSI sonde sensors at the RTC pier can estimated by its
measurement frequency as follows:

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Amount of data produced by the RTC's YSI sondes}
        \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}\hline 
        \textbf{YSIs} & \textbf{Rate} & \textbf{Hourly} & \textbf{Daily} &
        \textbf{Weeky} & \textbf{Monthly} & \textbf{Yearly}\\\hline 
        1 & 1 min & 3.04 Kb & 73.12 Kb & 511.87 Kb & 1.99 Mb & 23.99 Mb\\\hline 
        5 & 6 min & 15.23 Kb & 365.62 Kb & 2.5 Mb & 9.99 Mb & 119.97 Mb\\\hline 
        1 & 15 min & 0.5 Kb & 12.18 Kb & 85.31 Kb & 341.25 Kb & 3.99 Mb\\\hline 
        5 & 1 min & 2.54 Kb & 60.93 Kb & 426.56 Kb & 1.67 Mb & 19.99 Mb\\\hline
        1 & 6 min & 0.2 Kb & 4.87 Kb & 34.12 Kb & 136.5 Kb & 1.6 Mb\\\hline 
        5 & 15 min & 1.0 Kb & 24.37 Kb & 170.62 Kb & 682.5 Kb & 7.99 Mb\\\hline
        \end{tabular}
        \end{center}
\end{table}

As it is shown, the data produced by this type of sensor runs in the order of
at most hundreds of Megabytes a year, what can be considered a very low data
storage requirement. Upon collecting the data from sensors, the RTC lab staff
use automation scripts written in Matlab to process, index and distribute the
raw data in different formats, including the OPeNDAP \cite{opendap}, a format
widely used in the research institutions that promotes easier data exchange
among them.

Although the execution of the RTC SF-BEAM sensor network produces its
execution, many operational challenges were faced by this type of
infrastructure, when it comes to the data utilization of wireless sensors that
are deployed in offshore conditions, \cite{netbeams2009} described the
operational challenges faced by the RTC staff and proposed NetBEAMS, a
component-based infrastructure that uses common-off-the-shelfs (COTS) embedded
systems and software to automate the process of data collection. Next section
describes how NetBEAMS is used to extract data from the SF-BEAMS sensor network.

\section{NetBEAMS: a Real-World Case Study on Environmental Sensor Network}

The Networked Bay Environmental Assessment and Monitoring System, or NetBEAMS, 
is a work-in-progress joint project proposed by the department of Computer
Science and the RTC at San Francisco State University. As described by
\cite{netbeams2009}, the solution can be used to automated the data gathering
process of the SF-BEAMS as described in section \ref{sec:sfbeams}. This
section fully details how NetBEAMS is executed in order to collect data from
SF-BEAMS, describing pertinent design and architectural details.

One of the motivations for the inception of NetBEAMS was the automation
of the data gathering and processing from the SF-BEAMS sondes without requiring
human intervention. Most importantly, the decrease with the costs of constant 
offshore site visits by boat, where biologists need manually collect the data
from the sensor devices in order to take them for further processing at the
lab. For this reason, the NetBEAMS research group proposed the use of the Data
Sensor Platform (DSP) \cite{netbeams2009}, as the second attempt to automate
the operation of SF-BEAMS.

This following sections details the DSP, its architecture and how data is
transported from the network nodes to the main network sink, the RTC labs.

\subsection{The Data Sensor Platform (DSP)}

The Data Sensor Platform, in short DSP Platform, is based on a micro-kernel
architecture developed on top of the Java modular framework called OSGi
\cite{osgi}. The platform represents the foundation of the execution of
NetBEAMS, since it is executed from an embedded hardware called Gumstix
\cite{gumstix} using a Linux operating system and on top of the Java Virtual
Machine (JVM). In order to transport data from sensors hosts to the network
sink, the use of a cellular connection is used to transport the collected data
from the senso devices using the HTTP Protocol, as described by
\cite{netbeams2009}. Figure \ref{fig:sf-netbeams-node} shows the
architecture of the inhanced node designed by the NetBEAMS research group.

\begin{figure}
  \centering
  \includegraphics{../diagrams/DSP-Gateway-Node}
  \caption{The NetBEAMS Node enhancement with the Gumstix and the GSM Modem}
  \label{fig:sf-netbeams-node}
\end{figure}

The following list describes each of the devices in the NetBEAMS Node.

\begin{itemize}
  \item \textbf{YSI 6600EDS V2}: the YSI device is the regular sonde
  responsible for observing the environment. It is connected to the Gumstix
  embedded system, where the DSP is executed;
  \item \textbf{Gumstix console-vx}: it's a COTS ARM-based hardware that
  provides a computer-on-module environment for the development of small
  embedded systems. This is the main environment of data extraction and remote
  processing is accomplished. It runs a cross-compiled version of Linux Kernel
  2.6;
  \item \textbf{Huawei E220 USB Modem}: it is used to transfer the data from the
  Gumstix to the RTC Labs Data Center.
\end{itemize}

The DSP runs inside of the embedded system Gumstix. The following sections
details its design and process for extracting data.

\subsection{The DSP System Design}

This section focus on the development of a Software Platform for the NetBEAMS Gateway Embedded System, or on the system developed in the Gumstix device. As shown on Image 2.4.1.1, the NetBEAMS Gateway Node is an embedded system that contains a set of layer found in any computer. However, each of the resources described from the system were customized for the NetBEAMS system. The main components of such system can be summarized as follows:

    * Operating System: it uses a cross-compiled Gentoo Linux, which supports a wide variety of development tools, including the Java Virtual Machine (JVM), the underlying platform system;
    * Java Virtual Machine: it uses the JamVM, a ~200 KB version of the Sun Microsystems Java Virtual Machine, version 2.0;
    * OSGi Framework: it uses the Knopflerfish implementation of the OSGi 4.1 specification. More details in the following sections.
    * DSP Platform and other Bundles: The plug-and-play DSP Components are based on the OSGi bundles capabilities, which can reuse services registered in the OSGi framework.

>>>>

%\bibliography{thesis}  % uses the references stored in Chapter1Radar.bib

@@

\section{Sensor Networks Survey}

@@

\section{Persistence Storage for Sensor Networks}

@@

At first glance, the raw data serves as the primary product of the sensor
network. However, the first consideration that must take place is regarding the
purpose of the collected data, as \cite{sn-provenance} explains
differentiates between two main use of the sensor data:

@@

While real-time data stream are used and can potentially be discarded, the
archival data needs to be stored in a secondary memory. Nonetheless, given the
nature of the sensor device specification, the low-level data produced
by a given sensor device might not contain necessary information regarding the
identity of the data itself. The use of Data Provenance techniques is described
by \cite{sn-provenance} as an important step to provide the collected data 
specific metadata regarding its location and lineage.

@@

@@

\subsection{Query Processing for Storage Locations}
\label{sec:query-process}

@@

\begin{itemize}
  \item \textbf{In-Network Query Processing}: sensor nodes that provides
  real-time data stream usually replies to requests to properties. Likewise, when the
  data is used for archival purpose, the data can be located at a network node
  specialized in storing the neighbors collected data. In this way, when a
  query is issued to the sensor network, the query processor will
  collectively verify every storage node for the collected data.
  \item \textbf{Centralized Query Processing}: a main network sink receives the
  collected data from the peer nodes. In this case, the query processor queries
  the solo centralized node for the overall values of the participating sensor
  devices.
\end{itemize}

There are pros and cons related to both of the approaches of query processing,
which are also related to the infrastructure, sensor network nature and 
requirements. The main goal of the in-network query processing is to decrease
the amount of data transmitted from the sensor devices to the final data sink, as
sensor devices are usually limited by constraints such as energy utilization 
\cite{sn-storage03}. Along with this strategy, the literature describes
different proposals that aims at upgrading the Structured Query Language (SQL)
adding location-sensitive clauses to the language \cite{sn-db-newop}. However,
this main problem related to this approach is that it may potentially flood
all nodes with query messages \cite{sn-storage04}.

@@

A centralized network query processing processes data in a network node
specified as the network sink. The network sink usually is a network
node with a database server with the collection of data delivered by
all the sensor devices from the network. As a consequence of this
many-to-one communication strategy \cite{sn-storage02}, the query process does
not affect any of the sensor nodes in the network and, thus, maximizes the
sensor device utilization. Moreover, the use of the data for historical
purposes is easier to manage and give access to the primary end users of the
data, following the process of data processing, compression, aggregation, etc
\cite{sn-db-modeling02}.

@@

\section{Data Models and Query Engines for Sensor Networks}
\label{sec:data-models}

Different approaches are used to represent the collected data from sensors. The
use of simple data descriptors have been used in research groups which
expertise is not database systems, while others use such technology and the way
to represent data using more robust approaches. \cite{sn-data-model-survey}
have shown different data models and the advantages and desadvantages. The
following is a summary of these approaches used by the sensor network
community.

\begin{itemize}
  \item \textbf{Tabular Data Model}: comma-separeted values are very popular
  due to its simplicity. Although data is not index, it is an easy approach for data
 exchange among researchers.
  \item \textbf{Relational Data Model}: it is used in the majority of the
  projects. Most of them due to the easy access to database systems, with
  refactored versions that support the implementation of sensor networks and
  its requirements \cite{sn-db-tinydb};
  \item \textbf{Structured Data Model}: simiar to the relational model,
  structured data models are becoming popular due to the Internet. The use of
  XML data \cite{xml} is not yet popular, but there are a few projects using it
  \cite{sn-xml-usage01}\cite{sn-xml-usage01}, implementing its query by using a
  middleware \cite{sn-xml-middleware} \cite{sn-xml-query-engines}.
  \item \textbf{Key-Value Pair and Document-Oriented Data Model}: as
  an emerging data model, the Key-Value Pair (KVP) \cite{db-kvp} data model is
  becoming popular with the up-coming trend of cloud computing. Although there are no
  records of the use of this model data model in sensor networks, this work
  will investigate it.
\end{itemize}

Most importantly, the use of Data Provenance approaches to describe the
collected data is shown to be important \cite{sn-provenance}, given the nature
of the collected data. Therefore, this section describes these approaches for
modeling data collected from sensor devices, taking into account the need of
metadata for the data.

\subsection{Tabular Data Model and Query}

Simple use of Comma-Separated Values is an example os sensor network data. Data
is structured by the top header, and the values are in each line of the file.
The implementation of the SF-BEAMS \cite{sfbeams2006} uses this approach. In
summary, in a tabular data model

\begin{itemize}
  \item data is represented by columns in a flat file separated by a character
  that separates the values. One example of such data in sensor networks is the
  OPenDAP \cite{opendap} format, used by the SF-BEAMS project
  \cite{sfbeams2006}.
  \item data is queried using a text editor file, without any indexing
  associated with it. 
\end{itemize}

\subsection{Relational Data Model and Query}

Many different projects have used the relational model \cite{relational-model}
as the representation of data for sensor networks. In general, the following
properties apply to sensor networks:

\begin{itemize}
  \item data is structured in tables, and sensor properties are represented 
 by the columns of these tables. Data instances are represented by tuples or
 rows of these tables. Finally, table relationships are implemented based
 on the relational algebra abstraction, which is the mechanism used to 
describe the relationship among different data types.
  \item data is queried using the Structured Query Languange (SQL), and uses
  the definitions of the tables and columns to extract data.
\end{itemize}

Before the conception of such model, a database normalization process is
necessary to describe the data in an appropriate way. An example of such use
of the relational model is the TinyDB database systems \cite{sn-db-tinydb},
which uses the relational model to represent the types of sensors. Due to the
popularity of relational models, this is a popular choice since of the vast
availability of research on relational database models.

Although the implementation of a relational database seems to cover the needs
of data representation, \cite{sn-data-model-survey} described the problems
related to this approach. One such problem is that the relational model is not
suited for modeling time series data and metadata related to identify, time and
location. For the problems related to it, came the implementation of different
approaches that adds semantics to data types and different query clauses to the
Structured Query Language (SQL) to solve join \cite{sn-db-newop}.

\subsection{Structured Data Model and Query}

The use of XML is used to represent data in sensor networks has become popular
due the popularity of this data format. In general, the use of XML for
persistence can be summarized as follows:

\begin{itemize}
  \item data is structured in the collected data properties can be represented through the
XML \cite{xml} instances by using tags, having its model validated by the use
of XML Schema \cite{xml-schema}.
  \item in order to extract data from XML-based databases, the use of XML XPath
  is used to query XML documents.
\end{itemize}

Some examples of the application of XML in sensor networks in 
\cite{sn-xml-usage02} some have investigated the effecient use of XML in
sensor networks to represent the collected data \cite{sn-xml-usage01}. An
example of querying xml data in sensor networks described by
\cite{sn-xml-query-engines}, while others use XML as a means of data exchange
by implementing of middlewares to query and extract query data
\cite{sn-xml-middleware}. NetBEAMS is one of such sensor network that contains
an XML-based communication middleware for data exchange among the sensor nodes
\cite{netbeams2009}.

\subsection{Key-Value Pair and Document-Oriented Data Models}

The advance of distributed systems and the Internet have enabled the
development of powerful database systems that can be organized in the context
of database systems and how to represent models. An important up-coming
approach is the use of database systems that does not use structured language
for quering its data. One such example is called Key-Value Pair (KVP) data model
\cite{db-kvp}. Such model has the following properties:

\begin{itemize}
  \item data is structured in collections of key-value pairs, as it is done in
  hash data structure. The definition of the key is a given property with its
  associating value.
  \item the query process is using mechanics similar to programming or
  scripting language, that is, the use of APIs in a given programming language.
\end{itemize}

Different variants of such data model is the document-oriented data model,
where data is modelled using structured documents. However, these documents
have a dynamic structure that can be freely described without the use of an
schema validation mechanism as used in XML Schema. One example of such
approach is the use of the JSON data format \cite{json} in the implementation
of database systems which are becoming popular with the new trend called Cloud
Computing \cite{cloud-comp-architectures}. Databases implementing this strategy
is the mongoDB \cite{mongodb} and CouchDB \cite{mongodb}.

\subsection{Data Provenance for Sensor Networks}

Most of sensor networks implementations use the collected data from the sensor
devices after they observe the environment. As shown by
\cite{sn-data-model-survey}, different data models contains its pros and
cons to represent data. However, one important aspect in sensor networks is the
description of it collected data, as \cite{sn-provenance} discussed the
properties and nature of sensor networks generated data. First, the device's
sampling are usually in raw format or without obvious names. In order to
provide search over the collected sensor data, the data needs first to be
described and indexed, and this process can be related to Data Provenance
\cite{db-provenance}.

An example of Data Provenance application is the use of aggregation of data
over time to estimate the changing effects of the weather in a given region.
For this reason, the collected data needs descriptions or annotations that can
give enough information to track changes, that is, the history of how and when
a given data has changed. The following should be the foundation of a
provenance-aware collected data according to \cite{sn-provenance}:

While Data Provenance provides necessary mechanisms to describe sensor networks
data in general, others see the importance of the time space of sensor
networks. \cite{sn-time-series} describes strategies to used to cluster
distributed sensor data as time series of data.
